### 2017-10-04 Manuscript/replyToReferees.tex

\re{3. The 'Methods' section needs considerable restructuring. Preferably the toy (or simulated) example is pushed to the first part of the results section, and the methods section contains the derivation and structure of proposed methods, a brief description of Hartigan's dip test and a description of the used datasets.}

\smallskip
We have made particular effort to separate out the theoretical (methods) from the experimental (results). We find that in manuscripts that combine both, it is best to emphasize the difference, so that there is no doubt which is which, even at a glance. As such, we describe the used datasets where they are used, in the "Results - experiments" section. Further, in accordance with MIFlowCyt, the experimental details are summarized in the MIFlowCyt supplements and the data made freely available. We have added a comment to this effect at the beginning of the \emph{Methods} section and details of the data availability in the \emph{Supporting information}. We hope the Referee understands and accepts our choice of presentation.

\re{4. There is no description of the data that is used in this study whatsoever. It is not clear whether the data from ref. [1] were reused or acquired in the same way as described in ref. [1]. Both datasets cannot be accessed. Details about the data are not present, which they should be. Preferably, they should also be made available, please consider Flowrepository to freely store your flow cytometry data (especially for ref. [1], as this data can only accessed upon request from correponding author, although they are published in an open-access journal).}

\smallskip
We have addressed this, providing description of the data in accordance with MIFlowCyt. We have made the data and scripts which create the figures publicly available. We provide details in the \emph{Reply to Associate Editor}. Briefly, all data, scripts, and MIFlowCyt details can be found in
\href{https://github.com/AmirErez/BimodalLogspaceCytA}{https://github.com/AmirErez/BimodalLogspaceCytA}

@@

\smallskip
\re{5. The link to phase transitions needs more explanation or to be removed (line 53-54).}

\smallskip
Removed.

\re{6. The second method, how biological information can be used to properly estimate the mismatch between logarithmic and linear binning is in need of a proper presentation, as the idea seems an interesting one, but is not clearly laid out.}

\smallskip
To clarify this method, we have added a paragraph, which begins with "We briefly sketch what follows". (Last paragraph in page 12).
The purpose of this section is how, by considering both ERK1 and ppERK levels, one can ameliorate the mismatch between the two representations, resulting in more reliable gates.

@@

\re{
10. Please have a look at the manuscript organization of Cytometry part A: http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1552-4930/homepage/ForAuthors.html.
}

\smallskip
We have made the appropriate changes in manuscript organization.

\smallskip
\re{
Minor comments:
\\
11. There are quite a number of spelling mistakes and grammatical errors in the manuscript.}

\smallskip
We thank the editor for catching these mistakes, and hope that these errors did not obscure his or her first reading of the manuscript. We have taken great care in revising this version and are happy to work with the editor and / or representatives from Cytometry A to correct any remaining errors.

@@

\section*{Reply to Referee 2}

We thank the Referee for considering our work \re{"worth bringing to the attention of the readership of Cytometry A"} and that \re{"the illustration seems particularly enlightening"}. We also thank the Referee for the multiple constructive suggestions and agree with the prediction that this more terse manuscript conveys the message more readily. We detail each change following the Referee's comments below:

\smallskip
\re{The notion and terminology employed in the manuscript is inconsistent with standard usage in probability and statistics, which is unfortunate as it is likely to limit/inhibit readers. I appreciate that changing this might be time consuming, though I personally think it would be worthwhile.
}
\re{What is labeled as Q(I) and called a distribution in the present paper is instead a probability density function (pdf). Let us say there is a single continuous random variable, I, the intensity of a fluorescence then the pdf would usually labeled $f_I(x)$, with associated cumulative distribution function (cdf) being $F_I(x) = int_{-\infty}^x f_I(y) dy$. What is called P(log I) would then the pdf of the random variable log(I), which would normally be written as $f_{log(I)}(x)$.}

\smallskip
We thank the Referee for the suggestion regarding precise notation of the pdfs in the manuscript, and that the Referee acknowledges that such a change of notation is a time consuming. Certainly, for some of the readership of Cytometry A the suggested change of notation $Q(I)\rightarrow f_I(x)$ and $P(\log I)\rightarrow f_{\log I}(x)$ would improve readability. We fear, however, that a possibly comparable proportion of the readers might mistake $f_I$ and $f_{\log I}$ for having the same \emph{functional form}, just of a different argument, which is manifestly not the case in our example. Indeed, that is precisely the reason we chose $P$ and $Q$ to make this distinction explicit to the reader with a less formal mathematical background. More so, $P$, instead of $Q$, was granted to $f_{\log I}$ on merit that the reader is most likely acquainted with a logarithmic representation of the data, and is more habituated to a pdf denoted $P$. This choice of $P$ for the pdf of $\log I$ is consistent with Cytometry A's list of accepted abbreviations which chooses $P$ for probability.

### 2017-11-08 Manuscript/LogSpaceDistributionsCytometryA.bak.tex
After second round of review

\documentclass[11pt,a4paper,draft]{article}
\usepackage{setspace}
\usepackage{hyperref}
\doublespacing
%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig1.tif}
  \caption{\textbf{Log-normal mixture showing the mismatch in the number of peaks.} \textbf{Top row, red:} $P(\log I)$ vs. $\log I$ normalized to max. \textbf{Middle row, magenta:} $Q(I)$ vs. $I$ normalized to max, and plotted on a narrower range. \textbf{Bottom row, blue:} $Q(I)$ vs. $\log I$ normalized to max, rescaling the x-axis as in Ref.~\cite{Novo2008}. In the central column where $\sigma_1=\sigma_2=0.8$, $P(\log I)$ shows explicit bimodality whereas $Q(I)$ is unimodal. \textbf{Right column:} in this case, the variance $\sigma^2$ is large enough (see Eq.~\ref{eq:Plogy3Maxima}) that $P(\log I)$ has only one mode even though it is modeled as a mixture. In all cases: $I_1=100,\,I_2=1000,\,\alpha=0.5$ varying $\sigma_{1,2}=\{0.4,0.8,1.2\}$ (left to right columns, respectively). Leftmost column shows axes. }
  \label{fig:PlotsLinearLog}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig2.tif}
  \caption{\textbf{Graphical solution to count the number of extrema.} When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized. Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent when the red curve intersects $A$ at 3 points, whereas the blue curve only intersects $A$ once. In both cases, the loci of the extrema are different for the two distributions. The plots along the diagonal (which correspond to the cases in Fig.~\ref{fig:PlotsLinearLog}) show the case $\sigma_1=\sigma_2$ which simplifies $A(y)$ to a straight line (the general case being a parabola) in these axes.}
  \label{fig:ExtremaSolutions}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[width=1\linewidth]{Fig3.tif}
  \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario. }\textbf{(A)} Histogram of ppERK as plotted by FlowJo \cite{FlowJo}; \textbf{(B)} Histograms for $P(\log I)$ (red,dots) and $Q(I)$ (blue,dots) vs. $\log I$ as estimated from the data binned logarithmically. Note that plotting $Q(I)$ vs. $\log I$ is somewhat unusual but allows both $P$ and $Q$ to be plotted on the same axis. The red line shows the result of fitting $P(\log I)$ to a gaussian mixture model (Eq.~\ref{eq:LogNormalMix}), and the blue line is the estimate for $Q(I)$ from $P(\log I)$ according to Eq.~\ref{eq:P2Q}. The blue star indicates the location of the only maximum for $Q(I)$ obtained from Eq. \ref{eq:extremaQ}, despite the obvious two maxima in $P(\log I)$ (red). Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal. \textbf{(C)} Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
  \label{fig:Experiment}
\end{figure}
Given the prevalence and success of manual gating of FCM data in many situations, we wondered whether gating in logarithmic scale could be justified \emph{a posteriori}, based on biological knowledge. In the final part of our analysis, we aimed to include additional information such that the biological significance of the distributions in our single-cell measurements is better captured.

@@

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{Fig4.tif}
   \caption{\textbf{Analysis of experimental data with two correlated measurements.} \textbf{(A)} The joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$ as a heat map with its marginals plotted on the top and on its right. The correlation between ppERK and ERK1 levels is clear in the data. Dashed red (grey) lines are proposed manual gates according to the marginal (joint) distributions $P(\log I_{ppERK})$ $\left(P_2(\log I_{ppERK},\log I_{ERK1})\right)$. \textbf{(B)} Bayesian network depicted as a graphical model to show the flow of influence on the measurement of ppERK. The pair ERK1 and ppERK are in a template to suggest that there exist other pairs of correlated observables that depend on activation status and cell-to-cell variability.}
   \label{fig:ppERKvsERK1heatmap}
\end{figure}

\smallskip
\begin{figure}
 \centering
   \includegraphics[width=1\linewidth]{Fig5.tif}  
  \caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I_{ppERK})$ and $Q(I_{ppERK})$ vs. $\log I_{ppERK}$ (dots: data, lines: Gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{fig:NormppERKbyERK}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
 \includegraphics[clip,width=1\hsize]{Fig6.tif}
  \caption{\textbf{Testing Hartigan's p-value $p_u$ for unimodality on a weakly bimodal log-normal mixture.} The heat-map shows $p_u$ as a function of the number of bootstrap tests and number of events in each test, for $\sigma_1=0.4$ and $\sigma_2=0.8$ as in Fig.~\ref{fig:ExtremaSolutions}(top,middle). With less than $\sim10^5$ events, Hartigan's test for this case may misidentify the number of peaks.}
  \label{fig:CheckHartigan}
\end{figure}

\subsection*{Relation between ERK1 level and activation status}
\begin{figure}[!htb]
 \centering
   \includegraphics[width=1\linewidth]{Fig7.tif}  
  \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other. To quantify this difference, the inset shows the mutual information between $P(\log ERK1)$ and $P(\log ERK1 | Activation)$, with the circle pointing out the particular concentration of stimulus (out of all concentrations used in Ref.~\cite{Feinerman2008}) we chose to plot in this example. The chosen concentration has the highest mutual information, {\it i.e.}, the lowest ability to discern between the active and inactive states.}
  %\caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I)$ and $Q(I)$ vs. $\log I$ (dots: data, lines: gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{S1_Fig}
\end{figure}

>>>>

\documentclass[11pt,a4paper,final]{article}
\usepackage{setspace}
\usepackage{hyperref}
\doublespacing
%\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig1.png}
  \caption{\textbf{Log-normal mixture showing the mismatch in the number of peaks.} \textbf{Top row, red:} $P(\log I)$ vs. $\log I$ normalized to max. \textbf{Middle row, magenta:} $Q(I)$ vs. $I$ normalized to max, and plotted on a narrower range. \textbf{Bottom row, blue:} $Q(I)$ vs. $\log I$ normalized to max, rescaling the x-axis as in Ref.~\cite{Novo2008}. In the central column where $\sigma_1=\sigma_2=0.8$, $P(\log I)$ shows explicit bimodality whereas $Q(I)$ is unimodal. \textbf{Right column:} in this case, the variance $\sigma^2$ is large enough (see Eq.~\ref{eq:Plogy3Maxima}) that $P(\log I)$ has only one mode even though it is modeled as a mixture. In all cases: $I_1=100,\,I_2=1000,\,\alpha=0.5$ varying $\sigma_{1,2}=\{0.4,0.8,1.2\}$ (left to right columns, respectively). Leftmost column shows axes. }
  \label{fig:PlotsLinearLog}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[clip,width=1\hsize]{Fig2.png}
  \caption{\textbf{Graphical solution to count the number of extrema.} We test our peak counting method, based on the same equations as in Fig.~\ref{fig:PlotsLinearLog}, but with additional settings. When the red (blue) curves intersect the dashed black line, $P(\log I)$ ($Q(I)$) are extremized. Dashed black: $A(y)$, blue: $B_1(y)$ and red: $B_3(y)$. The mismatch between the number of extrema of $P(\log I)$ and $Q(I)$ is apparent when the red curve intersects $A$ at 3 points, whereas the blue curve only intersects $A$ once. In both cases, the loci of the extrema are different for the two distributions. The plots along the diagonal (which correspond to the cases in Fig.~\ref{fig:PlotsLinearLog}) show the case $\sigma_1=\sigma_2$ which simplifies $A(y)$ to a straight line (the general case being a parabola) in these axes.}
  \label{fig:ExtremaSolutions}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
  \includegraphics[width=1\linewidth]{Fig3.png}
  \caption{\textbf{Analysis of experimental data reveals the effect we describe in a real scenario. }\textbf{(A)} Histogram of ppERK as plotted by FlowJo \cite{FlowJo}; \textbf{(B)} Histograms for $P(\log I)$ (red,dots) and $Q(I)$ (blue,dots) vs. $\log I$ as estimated from the data binned logarithmically. Note that plotting $Q(I)$ vs. $\log I$ is somewhat unusual but allows both $P$ and $Q$ to be plotted on the same axis. The red line shows the result of fitting $P(\log I)$ to a gaussian mixture model (Eq.~\ref{eq:LogNormalMix}), and the blue line is the estimate for $Q(I)$ from $P(\log I)$ according to Eq.~\ref{eq:P2Q}. The blue star indicates the location of the only maximum for $Q(I)$ obtained from Eq. \ref{eq:extremaQ}, despite the obvious two maxima in $P(\log I)$ (red). Hartigan's unimodality p-values for $\log I$ (red) and $I$ (blue) are taken directly from the data without binning, corroborating that $\log I$ is bimodal whereas $I$ is unimodal. \textbf{(C)} Graphic solution of the extrema conditions as in Fig.~\ref{fig:ExtremaSolutions} explicitly reveals the three solutions for $P(\log I_*)$ (red line intersects black dashed) as opposed to the single solution for $Q(I_*)$ (the blue line intersects the black dashed line up beyond the plotted area, solution also plotted as blue star in the middle plot), indicating that $Q(I)$ has only one mode.}
  \label{fig:Experiment}
\end{figure}
Given the prevalence and success of manual gating of FCM data in many situations, we wondered whether gating in logarithmic scale could be justified \emph{a posteriori}, based on biological knowledge. In the final part of our analysis, we aimed to include additional information such that the biological significance of the distributions in our single-cell measurements is better captured.

@@

\begin{figure}
    \centering
    \includegraphics[width=1\linewidth]{Fig4.png}
   \caption{\textbf{Analysis of experimental data with two correlated measurements.} \textbf{(A)} The joint distribution $P_2(\log I_{ppERK},\log I_{ERK1})$ as a heat map with its marginals plotted on the top and on its right. The correlation between ppERK and ERK1 levels is clear in the data. Dashed red (grey) lines are proposed manual gates according to the marginal (joint) distributions $P(\log I_{ppERK})$ $\left(P_2(\log I_{ppERK},\log I_{ERK1})\right)$. \textbf{(B)} Bayesian network depicted as a graphical model to show the flow of influence on the measurement of ppERK. The pair ERK1 and ppERK are in a template to suggest that there exist other pairs of correlated observables that depend on activation status and cell-to-cell variability.}
   \label{fig:ppERKvsERK1heatmap}
\end{figure}

\smallskip
\begin{figure}
 \centering
   \includegraphics[width=1\linewidth]{Fig5.png}  
  \caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I_{ppERK})$ and $Q(I_{ppERK})$ vs. $\log I_{ppERK}$ (dots: data, lines: Gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{fig:NormppERKbyERK}
\end{figure}

@@

\begin{figure}[!ht]
 \centering
 \includegraphics[clip,width=1\hsize]{Fig6.png}
  \caption{\textbf{Testing Hartigan's p-value $p_u$ for unimodality on a weakly bimodal log-normal mixture.} The heat-map shows $p_u$ as a function of the number of bootstrap tests and number of events in each test, for $\sigma_1=0.4$ and $\sigma_2=0.8$ as in Fig.~\ref{fig:ExtremaSolutions}(top,middle). With less than $\sim10^5$ events, Hartigan's test for this case may misidentify the number of peaks.}
  \label{fig:CheckHartigan}
\end{figure}

\subsection*{Relation between ERK1 level and activation status}
\begin{figure}[!htb]
 \centering
   \includegraphics[width=1\linewidth]{Fig7.png}  
  \caption{{\bf Test for weak dependence of ERK1 and activation status.} Whereas independence implies that $P(\log ERK1) = P(\log ERK1 \vert Activation)$, in fact there is a weak dependence between them regardless of whether we employ a vertical (red, defined by threshold value $ppERK_*$) or a diagonal (grey) gate (the diagonal gate showing a weaker dependence); this is observed directly by noting that the different distributions in Fig.~\ref{S1_Fig} do not lie on top of each other. To quantify this difference, the inset shows the mutual information between $P(\log ERK1)$ and $P(\log ERK1 | Activation)$, with the circle pointing out the particular concentration of stimulus (out of all concentrations used in Ref.~\cite{Feinerman2008}) we chose to plot in this example. The chosen concentration has the highest mutual information, {\it i.e.}, the lowest ability to discern between the active and inactive states.}
  %\caption{\textbf{By dividing ppERK readings by ERK1, we can ameliorate the mismatch between the two representations.} \textbf{(A)} $P(\log I)$ and $Q(I)$ vs. $\log I$ (dots: data, lines: gaussian mixture fit) together with \textbf{(B)} their extrema analysis, showing that the second mode in log ppERK does not exist if the data is linearly binned. \textbf{(C)} The same treatment but for $\tilde{I} = I_{ppERK}/I_{ERK1}$, \textbf{(D)} shows that both $\tilde{I}$ and $\log \tilde{I}$ have two modes, thus normalizing ppERK levels by total ERK1 maintains the bimodal structure both in $P(\log \tilde{I})$ and in $Q(\tilde{I})$. }
  \label{S1_Fig}
\end{figure}
